Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-23 paper.qmd
Select the best representation for PBMC3k

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, 
                                                                 x = "UMAP1", y = "UMAP2"))) |>
  dplyr::mutate(cell_label = umap_pbmc$cell_label) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

@@

<!-- MSE plot to find which is the effective number of bins to use-->
```{r}
#| warning: false
#| echo: false
#| message: false

#tot_num_bin_vec <- 2:51
hex_size_vec <- seq(0.02, 2, by = 0.01)

vec <- stats::setNames(rep("", 7), c("num_bins", "error", "mse", "num_bins_x", "num_bins_y", "hex_size", "num_non_empty_bins"))  ## Define column names

mse_df_pbmc <- dplyr::bind_rows(vec)[0, ]
mse_df_pbmc <- mse_df_pbmc |>
  dplyr::mutate_if(is.character, as.numeric)

  num_bin_list <- calc_bins(data = umap_pbmc_scaled, 
            x = "UMAP1", y = "UMAP2", 
            hex_size = hex_size_vec[i], buffer_x = NA, buffer_y = NA)

  model_object <- fit_highd_model( training_data = training_data_pbmc, 
                                   nldr_df_with_id = umap_pbmc_scaled, 
                                   x = "UMAP1", y = "UMAP2", 
                                   num_bins_x = num_bins_x, 
                                   num_bins_y = num_bins_y, 
                                   x_start = NA, y_start = NA, 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_vec[i],
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "UMAP", 
                                   col_start_highd = "PC_")

  pred_emb_list <- predict_emb(test_data = training_data_pbmc, 
                                  df_bin_centroids = centroid_df_training, 
                                  df_bin = avg_df_training, type_NLDR = "UMAP")

  eval_list <- gen_summary(test_data = training_data_pbmc, 
                                  prediction_df = pred_df_training, 
                                  df_bin = avg_df_training, col_start = "PC_")

  mse_df_pbmc <- mse_df_pbmc |>
    tibble::add_row(num_bins = num_bins_x * num_bins_y,
                    error = eval_list$error,
                    mse = eval_list$mse,
                    num_bins_x = num_bins_x,
                    num_bins_y = num_bins_y,
                    hex_size = hex_size_vec[i],
                    num_non_empty_bins = NROW(centroid_df_training))

## If same total number of bins occurred only select ones with minimum error
### Obtain duplicate bins
dupli_bins <- mse_df_pbmc |> 
  dplyr::count(num_bins) |> 
  dplyr::filter(n > 1) |> 
  dplyr::pull(num_bins)

### Group split by duplicated bins
duplicate_df_list <- mse_df_pbmc |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_pbmc |>
  dplyr::filter(!(num_bins %in% dupli_bins))

### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_pbmc <- dplyr::bind_rows(not_dupli_df, duplicate_df)
```

mse_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  ylab("MSE") +
  xlab("total number of bins")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training PBMC3k dataset. What is the effective number of bins in UMAP to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 588, the slope of the MSE plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-msepltpbmc
#| fig-pos: H

mse_plot_pbmc
```

@@

```{r}
#| echo: false
#| fig-cap: "(a) Model in the 2D space overlaid on UMAP data, and (b) Error occurred by the high-D model for high-D data and the error mapped into 2D NLDR points."
#| label: fig-nldervisPBMCUMAP
#| fig-pos: H

@@

error_df <- error_df |>
  dplyr::mutate(type = paste0(n_neighbors, "_", min_dist))

abs_error_plot_pbmc <- ggplot(error_df, aes(x = num_non_empty_bins,
                                       y = error, group = type,
colour = type)) +
  geom_point() +
  geom_line() +
   # geom_vline(xintercept = 588, linetype="solid",
   #              color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  ylab("absolute error") +
  xlab("number of non-empty bins")

@@

>>>>

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, 
                                                                 x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

@@

@@

```{r}
#| echo: false
#| fig-cap: "(a) Model generated with 21 and 28 bins along x and y axes with hexagonal size 0.03 in the 2D space overlaid on UMAP data, and (b) high-D model error in model space."
#| label: fig-nldervisPBMCUMAP
#| fig-pos: H

@@

error_df <- error_df |>
  dplyr::mutate(type = paste0("n_neighbors: ", n_neighbors, ", min_dist: ", min_dist))

error_df_tsne <- read_csv("data/pbmc3k/error_df_tsne.csv", col_names = FALSE)
names(error_df_tsne) <- c("num_bins", "error", "mse", "num_bins_x", "num_bins_y", "hex_size", "num_non_empty_bins", "perplexity")
error_df_tsne <- error_df_tsne |>
  dplyr::mutate(type = paste0("perplexity: ", perplexity))

error_df <- bind_rows(error_df[, c(1:7, 10)], error_df_tsne[, c(1:7, 9)])

abs_error_plot_pbmc <- ggplot(error_df, aes(x = num_non_empty_bins,
                                       y = log(error), group = type,
colour = type)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 439, linetype="solid",
                color = "black", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7),
        legend.text = element_text(size = 5)) +
  scale_color_manual(values=c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a', '#b15928', '#ffff99')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")

```{r}
#| echo: false
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to 439, the lowest error occurred with the parameters perplexity: 51."
#| label: fig-abserrorPBMC
#| fig-pos: H
#| out-width: 100%

@@

tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_51.rds")
tsne_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tsne_pbmc, 
                                                                 x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tsne_pbmc))

tsne_plot_pbmc <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.5) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
        legend.key.height = unit(0.25, 'cm'),
        legend.key.width = unit(0.25, 'cm')) 
```

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from tSNE (perplexity = 51) applied for the PBMC3k dataset. Is this a best representation of the original data?"
#| label: fig-tsnesuggest
#| fig-pos: H

tsne_plot_pbmc 
```

<!--Fit the model-->

## Decide by looking at MSE plot
num_bins_x_pbmc <- 31
num_bins_y_pbmc <- 40
hex_size_pbmc <- 0.02

hb_obj_pbmc <- hex_binning(data = tsne_pbmc_scaled, x = "tSNE1", 
                      y = "tSNE2", num_bins_x = num_bins_x_pbmc, 
                      num_bins_y = num_bins_y_pbmc, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_pbmc, col_start = "tSNE")

tsne_data_with_hb_id_pbmc <- as.data.frame(do.call(cbind, hb_obj_pbmc$data_hb_id))

model_object <- fit_highd_model( training_data = training_data_pbmc, 
                                 nldr_df_with_id = tsne_pbmc_scaled, 
                                 x = "tSNE1", y = "tSNE2", 
                                 num_bins_x = num_bins_x_pbmc, 
                                 num_bins_y = num_bins_y_pbmc, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_pbmc,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "tSNE", 
                                 col_start_highd = "PC_")

benchmark_value_rm_lwd <- stats::quantile(df_bin_centroids_pbmc$std_counts,
                probs = c(0,0.25,0.5,0.75,1), names = FALSE)[2]

df_bin_centroids_low <- df_bin_centroids_pbmc |>
  dplyr::filter(std_counts <= benchmark_value_rm_lwd)

identify_rm_bins <- find_low_dens_hex(df_bin_centroids_all = df_bin_centroids_pbmc,
                                      num_bins_x = num_bins_x_pbmc,
                                      df_bin_centroids_low = df_bin_centroids_low)

df_bin_centroids_pbmc <- df_bin_centroids_pbmc |>
  dplyr::filter(!(hexID %in% identify_rm_bins))

## To find the benchmark value
benchmark_pbmc <- find_lg_benchmark(distance_edges = distance_pbmc, distance_col = "distance")
benchmark_pbmc <- 0.04

trimesh_removed_pbmc_tsne <- vis_rmlg_mesh(distance_edges = distance_pbmc, benchmark_value = benchmark_pbmc, tr_coord_df = tr_from_to_df_pbmc, distance_col = "distance")

trimesh_removed_pbmc_tsne <- trimesh_removed_pbmc_tsne +
  geom_point(data = tsne_pbmc_scaled, aes(x = tSNE1, y = tSNE2), alpha = 0.08) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

# df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), tsne_data_with_hb_id_pbmc)
# 
# show_langevitour(df_all_pbmc, df_bin_pbmc, df_bin_centroids_pbmc, benchmark_pbmc, distance_pbmc, "distance", col_start = "PC_")

df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), tsne_data_with_hb_id_pbmc)

error_df <- error_df |>
  mutate(type = if_else(total == 0, "no error",
                        if_else(total <= 5, "error 0-5",
                                if_else(total <= 10, "error 5-10",
                                        if_else(total <= 15, "error 10-15",
                                                if_else(total <= 20, "error 15-20",
                                                        "error greter than 20"))))))

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 0-5", "error 5-10", "error 10-15", "error 15-20", "error greter than 20"))) |>
    ggplot(aes(x = tSNE1,
               y = tSNE2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=rainbow_hcl(6)) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

```{r}
#| echo: false
#| fig-cap: "(a) Model generated with 31 and 40 bins along x and y axes with hexagonal size 0.02 in the 2D space overlaid on tSNE data, and (b) high-D model error in model space."
#| label: fig-nldervisPBMCtSNE
#| fig-pos: H

trimesh_removed_pbmc_tsne + plot_list2_pbmc_error +
plot_layout(guides='collect', ncol=2) &
  theme(legend.position='bottom') 
```
